---
title: "ModelScript_forClayton"
author: "Louis Jochems"
date: "2025-03-04"
output: html_document
---

```{r packages for riskmap model}
library(sdm)
library(dplyr)
library(plyr)
library(ggplot2)
library(terra)
library(leaflet)
options("sp_evolution_status" = 2)
library(sp)
library(sf)
library(caret)
library(pROC)

#setwd("Z:/Riskmap/ForClayton_2025")

sdm_data <- read.csv("Z:/path/to/RiskmapData_forClayton.csv")
#set path to dataframe 
```

Before running the model, I first had to filter out erroneous negative water depths and set slightly negative water depths (within -0.5 m margin of error) to zero. This is what the code does in the below chunk: 

```{r negative water depths }
neg_depths <- sdm_data %>%
 filter(Monthly_HD < 0)
#monthly_HD is the mean of monthly high water depths, which I 
#calculated by taking the mean of daily high water levels across each month 
#from the nearest gauge station to each point 
#and then subtracted the underlying elevation to get DEPTH 

# I used High water level vs. mean or low water level because I think that included
# more positive water depths

#HOWEVER 
#still a lot of depths within between -0.5 and zero. decided to keep and set those to zero  
hist(neg_depths$Monthly_HD, breaks = 28)

#filter out any data more -0.5 
sdm_data <- sdm_data %>% 
  filter(Monthly_HD > -0.5)


#function to set all "slightly" negative depths to 0 
pos_func <- function(x) {
  x$Monthly_HD  <- ifelse(x$Monthly_HD < 0, 0, x$Monthly_HD )
  x$Monthly_MD <- ifelse(x$Monthly_MD < 0, 0, x$Monthly_MD)
  x$Monthly_LD <- ifelse(x$Monthly_LD < 0, 0, x$Monthly_LD)
  x$Daily_D <- ifelse(x$Daily_D < 0, 0, x$Daily_D)
  return(x)
}

sdm_data <- pos_func(sdm_data) 

hist(sdm_data$Monthly_HD)
#no more negative depths 
```

Next, I need to remove all observations with NA values for NDVI. These ended up being most if not all points from 2015. At the time I hadn't checked in GEE whether they had real values. I guess there was a gap in available imagery from Sentinel back then...

```{r remove NA values for NDVI}
sdm_data16_23 <- sdm_data %>%
  filter(!is.na(medianNDVI))

range(sdm_data16_23$medianNDVI)
range(sdm_data16_23$year)
#no more NAs and no more points from 2015. 
```

Now onto running models! For context here are the names/explanation of the model variables:

'hyd_bin' = "Hydrocharis binary" 0: EFB Absence, 1: EFB Presence 
'MeanFetch' = Mean fetch values across all years available. Extracted from nearest pixel to each point 
'Dist_NearBL' = Distance to nearest boat launch for each point (in m)
'Dist_NearEFB' = Distance to nearest known EFB presence point (in m)
'medianVH' = Median monthly* VH polarization value across all Sent 1 images for intersecting pixel to each point 
'medianVV' = Median monthly VH polarization ""
'medianVH_VV_ratio' = Median monthly VH_VV ratio "" 
'medianNDVI' = Median monthly NDVI value across all Sent 2 images for intersecting pixel to each point 
'medianNDWI' = Median monthly NDWI "" 

*monthly values calculated across images of month in which observation was made 

```{r model with ALL predictors}
#grid object for variables to include in random forest trees 
#9 predictor variables total 
tunegrid <- expand.grid(.mtry = (1:9)) 


caret_all <- train(hyd_bin ~ MeanFetch + Dist_NearBL + Dist_NearEFB + 
                   medianVH + medianVV + medianVH_VV_ratio + 
                   medianNDVI + medianNDWI + Monthly_HD, 
                   data = sdm_data16_23,
                   method = 'rf',
                   tuneGrid = tunegrid, # 1:9, range of # of variables to include for trees 
                   trControl = trainControl(method = 'repeatedcv', # Use cross-validation
                                            number = 5,# 5 folds 
                                            repeats = 50, # 50 model runs 
                                            classProbs = T,
                                            savePredictions = T)
                   )
#model runs, but while it does you get a warning about only two outcomes for response
#and whether you want a RF classification vs. regression... 
#I think we want regression so that we can get probabilities for EFB habitat 
#assigned to each pixel, rather than a binary yes/no for predicting EFB habitat 

#for different model metrics 
summary(caret_all)

#save .RData model object so that we can apply across entire raster composites 
save(caret_all, file = "/bsuhome/path/to/CaretM_AllPred.RData")

```

